LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-11-07 main.tex
Switch to two column layout
To detect overfull boxes use:
\documentclass[5p,draft]{elsarticle}
\usepackage[final]{listings}

Additional hints: http://tex.stackexchange.com/a/2780/29072

% FIXME Add 5p option to get JCompSci two column layout
\documentclass{elsarticle}

@@

\definecolor{gray}{rgb}{0.9,0.9,0.9}

\lstloadlanguages{Matlab}
\lstset{
  basicstyle=\small,
  breaklines=true,
  frame=single,
  float,
  columns=fullflexible,
  numbers=left,
  stepnumber=2,
  backgroundcolor=\color{gray},
  escapeinside={\%@}{\^^M}
}

@@

% FIXME Reviewer #5, comment #5: Please provide information regarding practical applications of this algorithmic software tool.
Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.

@@

Here we present an algorithmic introduction to geodesic LCS detection.
% FIXME JCompSci editor: Replace GitHub link with http://runmycode.org/
% FIXME Reviewer #3, comment #4: Add documentation within the github repository to show some reference to the APIs and how users may start to run a his/her own application.
% FIXME Reviewer #4: Please make sure the code is downloadable once the paper is published
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is available at: http://github.com/jeixav/LCS-Tool}. This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present form demonstration scripts distributed with LCS Tool.

Other software for LCS detection exists.
ManGen\citep{lekien03:_time} calculates the FTLE and advects material curves in two-dimensional velocity fields.
It includes a graphical user interface and uses the Message Passing Interface standard for parallel calculations.
Newman\citep{toit10:_trans} calculates the FTLE using dimension independent code.
It assists ridge extraction of FTLE fields and supports analytic and dataset based velocity definitions.
FlowTK\citep{ameli14:_devel_effic_flexib_pipel_lagran} calculates the FTLE in two and three-dimensions on Cartesian and unstructured grids.
The NVIDIA CUDA parallel computing platform is used for fast computations and Kitware's ParaView data analysis and visualization application is used for a user interface.

These packages target the automated generation of FTLE plots to aid the visual assessment of hyperbolic LCSs.
LCS Tool also has functions to generate FTLE plots, but its emphasis is to provide geodesic extraction of LCSs as parametrized material curves, and extend the scope of such extraction to elliptic LCSs.

@@

We seek positions of closed material lines at time $t_0$ that prevail as coherent Lagrangian vortex boundaries (or \emph{elliptic LCSs}) over a time interval $[t_0,t]\subset[t_-,t_+]$. \citet{haller13:_coher_lagran} argue that such initial material line positions are closed stationary curves of the averaged strain functional
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^{\sigma} \frac{\sqrt{\langle r'(s),C_{t_0}^t(r(s))r'(s)\rangle}}{\sqrt{\langle r'(s),r'(s)\rangle}} \dif s,
\]
obtained by averaging the tangential strain arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma].$
Solutions to this variational problem turn out to be closed orbits of one of two parametrized vector-field families
\begin{equation}
\eta_\pm^\lambda = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}}\xi_2,
\label{eq:eta}
\end{equation}
with $\lambda > 0$ playing the role of a parameter. Such closed orbits satisfy the differential equation
\begin{equation}
r' = \eta_\pm^\lambda(r),
\label{eq:etafields}
\end{equation}
which coincide with null-geodesics of the Lorentzian metric family 
\[
e_\lambda(u,v) = \left\langle u,\left[D_{t_0}^t(r) - \lambda^2 I\right] v \right\rangle.
\]
For this reason, we refer to the computation of elliptic LCSs as limit cycles of \cref{eq:etafields} as \emph{geodesic detection of elliptic LCSs}.

@@

Shrinklines and stretchlines coincide with the null-geodesics of the Lorentzian metric $h(u,v) = \left\langle u,D_{t_0}^t(r) v \right\rangle$. For this reason, we refer to the computation of hyperbolic LCSs as strongest normally-repelling or normally-attracting orbits of \cref{eq:etafields} as \emph{geodesic detection of hyperbolic LCSs}.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\tabularnewline
\hline
\lstinline!derivative! & function handle for flow velocity equations\tabularnewline
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\tabularnewline
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\tabularnewline
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default: \lstinline!true!.\tabularnewline
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default: \lstinline!false!.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\tabularnewline
\hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

@@

The auxiliary grid (cf. \cref{t:Cauchy-Green algorithm}) comprises four points placed symmetrically around each point of the Cartesian grid (\cref{f:main and auxiliary grids}). These points are used to achieve increased accuracy in the finite-difference approximation
\[
\nabla F_{t_0}^t(x_0) \approx
\left(\begin{array}{cc}
\frac{x_1(t;t_0,x_0 + \delta x_1) - x_1(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_1(t;t_0,x_0+\delta x_2) - x_1(t;t_0,x_0-\delta x_2)}{2\vert\delta x_2\vert}\\
\frac{x_2(t;t_0,x_0+\delta x_1) - x_2(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_2(t;t_0,x_0 + \delta x_2) - x_2(t;t_0,x_0 - \delta x_2)}{2\vert\delta x_2\vert}
\end{array}\right)
\]
of $\nabla F_{t_0}^t$ at a point $x_0$ of the Cartesian grid. Here $\delta x_i$ is a vector of length $\vert\delta x_i\vert > 0$ that points from the Cartesian grid-point $x_0$ in the $i^\text{th}$ coordinate direction (\cref{f:main and auxiliary grids}). Computational improvements arising from the use of the auxiliary grid over simply using the nearest points of the main grid were reported in \citet{farazmand12:_comput_lagran}. Experience suggests setting the auxiliary grid spacing to 1-10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of the main grid (filled circles) and the auxiliary grid (empty circles) used in the computation of the derivative of the flow map in \cref{eq:CG} of the Cauchy-Green strain tensor. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

The function \lstinline!eig_cgStrain! of LCS Tool provides the option to calculate Cauchy-Green eigenvectors from the auxiliary grid using eigenvalues calculated from the main grid. We have found that for flows defined analytically, the eigenvalues can reliably be calculated from the main grid. For flows defined by datasets, using the auxiliary grid for eigenvalue calculations gives better results.

As mentioned above, a typical main grid for the Cauchy-Green strain tensor comprises $500 \times 500$ points. This means that after the addition of 4 auxiliary grid points around each main grid point, \cref{eq:velocity} must be integrated over 1.25 million initial conditions.

@@

The Cauchy-Green strain tensor generically admits point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\citep{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

Velocity fields defined by datasets require pre-processing before they are used in the numerical integration of \cref{eq:velocity}. This requires spatial and temporal interpolation that enables the evaluation of the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. In \cref{sec:oceandataset}, we present an ocean dataset example with details of possible interpolation functions.

@@

% FIXME Need to update table to account for lambda range
\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax for LCS Tool elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{graphics/variable_step_integration}
\end{center}
\caption{Schematic illustration of the variable-time-step $\lambda$-line integration. At the initial point, there is an orientation discontinuity at the lower-right grid point that must be corrected prior to linear interpolation. At point \#1, no orientation discontinuities are present. At point \#2 all interpolated $\eta_\pm^\lambda$ vectors must be rotated by 180° to match the orientation of the trajectory.}
\label{f:variable step integration}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare map obtained for an elliptic LCS. The abscissa, $s$, represents distance along Poincare section. The ordinate, $p(s) - s$, represents distance of Poincare orbit return point from initial position. Circle markers indicate closed orbit positions. The filled circle indicates the outermost fixed point of the Poincare map, marking the intersection of a coherent Lagrangian cortex boundary with the Poincare section.}
\label{f:Poincare return map}
\end{figure}

@@

As discussed in \cref{sec:Hyperbolic LCSs}, positions of hyperbolic LCSs at time $t_0$ are found as the strongest repelling orbits of the vector field \cref{eq:shrinkline} (repelling LCSs), and strongest attracting orbits of the vector field \cref{eq:stretchline} (attracting LCSs). By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$. We identify the strongest repelling shrinklines as those crossing a local maximum of the $\lambda_2(x_0)$ field. Similarly, we identify the strongest attracting stretchlines as those crossing a local minimum of the $\lambda_1(x_0)$ field. These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of saddle points to the present finite-time, temporally aperiodic flow setting.

The main steps of our hyperbolic LCS detection algorithm are enumerated in \cref{t:Hyperbolic LCS algorithm}. The basic function to compute hyperbolic LCSs in LCS Tool is \lstinline!seed_curves_from_lambda_max! and its syntax is given in \cref{t:seed_curves_from_lambda_max syntax}.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![curvePosition,curveInitialPosition] = seed_curves_from_lambda_max(distance,cgEigenvalue,cgEigenvector,flowDomain,flowResolution)!}\tabularnewline
\hline
\lstinline!distance! & threshold distance for placement of $\lambda_2(x_0)$ maxima\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!flowDomain! & $2 \times 2$ array to define flow domain\tabularnewline
\hline
\lstinline!flowResolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\lstinline!nMaxCurves! & optional maximum number of curves (i.e. shrinklines of stretchlines) to generate. Default: \lstinline!numel(cgEigenvalue)!.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{Syntax of the function \lstinline!seed_curves_from_lambda_max!}
\label{t:seed_curves_from_lambda_max syntax}
\end{table}

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_coherent_eddy}
\includegraphics[width=.49\textwidth]{graphics/ocean_dataset/LCS_fwd_colortracer}
\caption{(Left) Elliptic LCS (green) with repelling LCSs (red) continuing into its interior. (Right) Final positions $x(t) = (x_1(t),x_2(t)) = F_{t_0}^t(x_0)$ of tracer particles. $x_1(t)$ position encoded in red and $x_2(t)$ position encoded in green. Note the continued impact of the repelling hyperbolic LCSs (black) on tracer patterns inside the elliptic LCS (white).}
\label{f:ocean_dataset_colortracer}
\end{figure}

@@

\begin{figure}[hbt]
\centering
\includegraphics[width=.8\textwidth]{graphics/double_gyre/lambda_lcs_convergence}\caption{Convergence of closed $\lambda$-lines for increasing main-grid resolution in the double-gyre model. The outermost closed $\lambda$-line (bold green line) is the vortex boundary. (Top) {[}500x250{]}, $\lambda$-values
of closed orbits from interior to exterior, left gyre: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right gyre: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04)$; (Middle) {[}750x375{]}, left: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)$; (Bottom) {[}1000x500{]}, left: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)$.}
\label{f:double gyre lambda LCS convergence}
\end{figure}

\Cref{f:double gyre lambda hyperbolic LCS} shows the complete picture of elliptic and hyperbolic LCSs of the double gyre at this resolution. \Cref{l:double gyre lambda strain stretch LCS} lists the corresponding LCS Tool commands. The maximal length of shrinklines and stretchlines, \lstinline!shrinklineMaxLength! and \lstinline!stretchlineMaxLength!, is set to $20$, a multiple of the domain size, since the hyperbolic LCSs may wind around vortices several times.

>>>>

\documentclass[5p]{elsarticle}

@@

\definecolor{grey}{rgb}{0.9,0.9,0.9}

\lstloadlanguages{Matlab}
\lstset{
  basicstyle=\small,
  breaklines=true,
  frame=single,
  float,
  columns=fullflexible,
  numbers=left,
  stepnumber=2,
  backgroundcolor=\color{grey},
  escapeinside={\%@}{\^^M}
}

@@

\begin{sloppypar}
% FIXME Reviewer #5, comment #5: Please provide information regarding practical applications of this algorithmic software tool.
Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.
\end{sloppypar}

@@

\begin{sloppypar}
Here we present an algorithmic introduction to geodesic LCS detection.
% FIXME JCompSci editor: Replace GitHub link with http://runmycode.org/
% FIXME Reviewer #3, comment #4: Add documentation within the github repository to show some reference to the APIs and how users may start to run a his/her own application.
% FIXME Reviewer #4: Please make sure the code is downloadable once the paper is published
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is available at: http://github.com/jeixav/LCS-Tool}. This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present form demonstration scripts distributed with LCS Tool.
\end{sloppypar}

\begin{sloppypar}
Other software for LCS detection exists.
ManGen\citep{lekien03:_time} calculates the FTLE and advects material curves in two-dimensional velocity fields.
It includes a graphical user interface and uses the Message Passing Interface standard for parallel calculations.
Newman\citep{toit10:_trans} calculates the FTLE using dimension independent code.
It assists ridge extraction of FTLE fields and supports analytic and dataset based velocity definitions.
FlowTK\citep{ameli14:_devel_effic_flexib_pipel_lagran} calculates the FTLE in two and three-dimensions on Cartesian and unstructured grids.
The NVIDIA CUDA parallel computing platform is used for fast computations and Kitware's ParaView data analysis and visualization application is used for a user interface.
\end{sloppypar}

\begin{sloppypar}
These packages target the automated generation of FTLE plots to aid the visual assessment of hyperbolic LCSs.
LCS Tool also has functions to generate FTLE plots, but its emphasis is to provide geodesic extraction of LCSs as parametrized material curves, and extend the scope of such extraction to elliptic LCSs.
\end{sloppypar}

@@

\begin{sloppypar}
We seek positions of closed material lines at time $t_0$ that prevail as coherent Lagrangian vortex boundaries (or \emph{elliptic LCSs}) over a time interval $[t_0,t]\subset[t_-,t_+]$. \citet{haller13:_coher_lagran} argue that such initial material line positions are closed stationary curves of the averaged strain functional
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^{\sigma} \frac{\sqrt{\langle r'(s),C_{t_0}^t(r(s))r'(s)\rangle}}{\sqrt{\langle r'(s),r'(s)\rangle}} \dif s,
\]
obtained by averaging the tangential strain arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma].$
Solutions to this variational problem turn out to be closed orbits of one of two parametrized vector-field families
\begin{equation}
\eta_\pm^\lambda = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}}\xi_2,
\label{eq:eta}
\end{equation}
with $\lambda > 0$ playing the role of a parameter. Such closed orbits satisfy the differential equation
\begin{equation}
r' = \eta_\pm^\lambda(r),
\label{eq:etafields}
\end{equation}
which coincide with null-geodesics of the Lorentzian metric family 
\[
e_\lambda(u,v) = \left\langle u,\left[D_{t_0}^t(r) - \lambda^2 I\right] v \right\rangle.
\]
For this reason, we refer to the computation of elliptic LCSs as limit cycles of \cref{eq:etafields} as \emph{geodesic detection of elliptic LCSs}.
\end{sloppypar}

@@

\begin{sloppypar}
Shrinklines and stretchlines coincide with the null-geodesics of the Lorentzian metric $h(u,v) = \left\langle u,D_{t_0}^t(r) v \right\rangle$. For this reason, we refer to the computation of hyperbolic LCSs as strongest normally-repelling or normally-attracting orbits of \cref{eq:etafields} as \emph{geodesic detection of hyperbolic LCSs}.
\end{sloppypar}

@@

\begin{table*}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{.95\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\tabularnewline
\hline
\lstinline!derivative! & function handle for flow velocity equations\tabularnewline
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\tabularnewline
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\tabularnewline
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default: \lstinline!true!.\tabularnewline
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default: \lstinline!false!.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\tabularnewline
\hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table*}

@@

The auxiliary grid (cf. \cref{t:Cauchy-Green algorithm}) comprises four points placed symmetrically around each point of the Cartesian grid (\cref{f:main and auxiliary grids}). These points are used to achieve increased accuracy in the finite-difference approximation
\begin{gather*}
\nabla F_{t_0}^t(x_0) \approx
\left(\begin{array}{cc}
\alpha_{11} & \alpha_{12}
\\
\alpha_{21} & \alpha_{22}
\end{array}\right)
\\
\alpha_{ij} \equiv \frac{x_i(t;t_0,x_0+\delta x_j) - x_i(t;t_0,x_0-\delta x_j)}{2\vert\delta x_j\vert}
\end{gather*}
of $\nabla F_{t_0}^t$ at a point $x_0$ of the Cartesian grid. Here $\delta x_j$ is a vector of length $\vert\delta x_j\vert > 0$ that points from the Cartesian grid-point $x_0$ in the $j^\text{th}$ coordinate direction (\cref{f:main and auxiliary grids}). Computational improvements arising from the use of the auxiliary grid over simply using the nearest points of the main grid were reported in \citet{farazmand12:_comput_lagran}. Experience suggests setting the auxiliary grid spacing to 1-10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=.475\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of the main grid (filled circles) and the auxiliary grid (empty circles) used in the computation of the derivative of the flow map in \cref{eq:CG} of the Cauchy-Green strain tensor. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

\begin{sloppypar}
The function \lstinline!eig_cgStrain! of LCS Tool provides the option to calculate Cauchy-Green eigenvectors from the auxiliary grid using eigenvalues calculated from the main grid. We have found that for flows defined analytically, the eigenvalues can reliably be calculated from the main grid. For flows defined by datasets, using the auxiliary grid for eigenvalue calculations gives better results.
\end{sloppypar}

\begin{sloppypar}
As mentioned above, a typical main grid for the Cauchy-Green strain tensor comprises $500 \times 500$ points. This means that after the addition of 4 auxiliary grid points around each main grid point, \cref{eq:velocity} must be integrated over 1.25 million initial conditions.
\end{sloppypar}

@@

\begin{sloppypar}
The Cauchy-Green strain tensor generically admits point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\citep{delmarcelle94}, and hence lie off the computational grid with probability one.
\end{sloppypar}

@@

Velocity fields defined by datasets require preprocessing before they are used in the numerical integration of \cref{eq:velocity}. This requires spatial and temporal interpolation that enables the evaluation of the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. In \cref{sec:oceandataset}, we present an ocean dataset example with details of possible interpolation functions.

@@

% FIXME Need to update table to account for lambda range
\begin{table*}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{.95\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in \cref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{.95\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax for LCS Tool elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.475\textwidth]{graphics/variable_step_integration}
\end{center}
\caption{Schematic illustration of the variable-time-step $\lambda$-line integration. At the initial point, there is an orientation discontinuity at the lower-right grid point that must be corrected prior to linear interpolation. At point \#1, no orientation discontinuities are present. At point \#2 all interpolated $\eta_\pm^\lambda$ vectors must be rotated by 180° to match the orientation of the trajectory.}
\label{f:variable step integration}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.475\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare map obtained for an elliptic LCS. The abscissa, $s$, represents distance along Poincare section. The ordinate, $p(s) - s$, represents distance of Poincare orbit return point from initial position. Circle markers indicate closed orbit positions. The filled circle indicates the outermost fixed point of the Poincare map, marking the intersection of a coherent Lagrangian cortex boundary with the Poincare section.}
\label{f:Poincare return map}
\end{figure}

@@

\begin{sloppypar}
As discussed in \cref{sec:Hyperbolic LCSs}, positions of hyperbolic LCSs at time $t_0$ are found as the strongest repelling orbits of the vector field \cref{eq:shrinkline} (repelling LCSs), and strongest attracting orbits of the vector field \cref{eq:stretchline} (attracting LCSs). By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$. We identify the strongest repelling shrinklines as those crossing a local maximum of the $\lambda_2(x_0)$ field. Similarly, we identify the strongest attracting stretchlines as those crossing a local minimum of the $\lambda_1(x_0)$ field. These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of saddle points to the present finite-time, temporally aperiodic flow setting.
\end{sloppypar}

\begin{sloppypar}
The main steps of our hyperbolic LCS detection algorithm are enumerated in \cref{t:Hyperbolic LCS algorithm}. The basic function to compute hyperbolic LCSs in LCS Tool is \lstinline!seed_curves_from_lambda_max! and its syntax is given in \cref{t:seed_curves_from_lambda_max syntax}.
\end{sloppypar}

@@

\begin{table*}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{.95\textwidth}|}
{\lstinline![curvePosition,curveInitialPosition] = seed_curves_from_lambda_max(distance,cgEigenvalue,cgEigenvector,flowDomain,flowResolution)!}\tabularnewline
\hline
\lstinline!distance! & threshold distance for placement of $\lambda_2(x_0)$ maxima\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!flowDomain! & $2 \times 2$ array to define flow domain\tabularnewline
\hline
\lstinline!flowResolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default: \lstinline![false,false]!.\tabularnewline
\hline
\lstinline!nMaxCurves! & optional maximum number of curves (i.e. shrinklines of stretchlines) to generate. Default: \lstinline!numel(cgEigenvalue)!.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{Syntax of the function \lstinline!seed_curves_from_lambda_max!}
\label{t:seed_curves_from_lambda_max syntax}
\end{table*}

@@

\begin{figure}
\centering
\includegraphics[width=.475\textwidth]{graphics/ocean_dataset/LCS_fwd_coherent_eddy}
\includegraphics[width=.475\textwidth]{graphics/ocean_dataset/LCS_fwd_colortracer}
\caption{(Top) Elliptic LCS (green) with repelling LCSs (red) continuing into its interior. (Bottom) Final positions $x(t) = (x_1(t),x_2(t)) = F_{t_0}^t(x_0)$ of tracer particles. $x_1(t)$ position encoded in red and $x_2(t)$ position encoded in green. Note the continued impact of the repelling hyperbolic LCSs (black) on tracer patterns inside the elliptic LCS (white).}
\label{f:ocean_dataset_colortracer}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.475\textwidth]{graphics/double_gyre/lambda_lcs_convergence}\caption{Convergence of closed $\lambda$-lines for increasing main-grid resolution in the double-gyre model. The outermost closed $\lambda$-line (bold green line) is the vortex boundary. (Top) {[}500x250{]}, $\lambda$-values
of closed orbits from interior to exterior, left gyre: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right gyre: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04)$; (Middle) {[}750x375{]}, left: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)$; (Bottom) {[}1000x500{]}, left: $\lambda=(0.97, 0.98, 0.99, 1.00, 1.01)$, right: $\lambda=(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)$.}
\label{f:double gyre lambda LCS convergence}
\end{figure}
